---
title: "Graphics and Data <br /> Visualization in R"
author: Thomas Girke
date: May 20, 2021
output: 
  ioslides_presentation:
    keep_md: yes
    logo: ./images/ucr_logo.png
    widescreen: yes
    df_print: paged
    smaller: true
subtitle: "Overview of Graphics Environments in R" 
bibliography: bibtex.bib
---

<!---
- ioslides manual: 
   https://bookdown.org/yihui/rmarkdown/ioslides-presentation.html

- Compile from command-line
Rscript -e "rmarkdown::render('rgraphics_slides.Rmd'); knitr::knit('rgraphics_slides.Rmd', tangle=TRUE)"
-->

<!---
  Note: following css chunks are required for scrolling support beyond slide boundaries
-->

<style>
slides > slide {
  overflow-x: auto !important;
  overflow-y: auto !important;
}
</style>

```{css, echo=FALSE}
pre {
  max-height: 300px;
  overflow-y: auto;
}

pre[class] {
  max-height: 300px;
}
```

```{css, echo=FALSE}
.scroll-300 {
  max-height: 300px;
  overflow-y: auto;
  background-color: inherit;
}
```

# Outline

- <div class="white">__Overview__</div>
- Base Graphics
- Grid Graphics
- Graphics with _lattice_
- Graphics with _ggplot2_ (_ggplotly_) 
- Specialty Graphics
- Genome Graphics
- References

## Overview

### Graphics in R

-   Powerful environment for visualizing data
-   Integrated graphics and statistics infrastructure
-   Publication quality graphics
-   Fully programmable
-   Highly reproducible
-   Full [LaTeX](http://www.latex-project.org/), [Sweave](http://www.stat.auckland.ac.nz/~dscott/782/Sweave-manual-20060104.pdf), [knitr](http://yihui.name/knitr/) and [R Markdown](http://rmarkdown.rstudio.com/) support.
-   Many R packages with graphics utilities

### Documentation on Graphics in R

- General 
    - [Graphics Task Page](http://cran.r-project.org/web/views/Graphics.html)
    - [R Graph Gallery](http://www.r-graph-gallery.com/)
    - [R Graphical Manual](https://www.imsbio.co.jp/RGM/R_image_list?page=575&init=true)
    - [Paul Murrell’s book R (Grid) Graphics](http://www.stat.auckland.ac.nz/~paul/RGraphics/rgraphics.html)

- Interactive graphics
    - [rggobi (GGobi)](http://www.ggobi.org/)
    - [iplots](http://www.iplots.org/)
    - [Open GL (`rgl`)](http://rgl.neoscientists.org/gallery.shtml)
    - [plotly](https://plotly.com/r/)

### Graphics Environments

- Viewing and savings graphics in R
    - On-screen graphics
    - postscript, pdf, svg
    - jpeg/png/wmf/tiff/...

- Four major graphics environments
    - Low-level infrastructure
        - R Base Graphics (low- and high-level)
        - `grid`: [Manual](http://www.stat.auckland.ac.nz/~paul/grid/grid.html), [Book](http://www.stat.auckland.ac.nz/~paul/RGraphics/rgraphics.html)
    - High-level infrastructure
        - `lattice`: [Manual](http://lmdvr.r-forge.r-project.org), [Intro](http://www.his.sunderland.ac.uk/~cs0her/Statistics/UsingLatticeGraphicsInR.htm), [Book](http://www.amazon.com/Lattice-Multivariate-Data-Visualization-Use/dp/0387759689)
        - `ggplot2`: [Manual](http://docs.ggplot2.org/current/), [Intro](http://www.ling.upenn.edu/~joseff/rstudy/summer2010_ggplot2_intro.html), [Book](http://had.co.nz/ggplot2/book/)

## `ggplot2`

### `ggplot`: scatter plots

#### Basic example

```{r ggplot_scatter_plot1, eval=TRUE}
library(ggplot2)
set.seed(1410)  
dsmall <- as.data.frame(diamonds[sample(nrow(diamonds), 1000), ])
p <- ggplot(dsmall, aes(carat, price, color=color)) + 
            geom_point(size=4)
print(p) 
```

Interactive version of above plot can be generated with the `ggplotly` function from 
the `plotly` package.

```{r ggplot_scatter_plot_interactive1, eval=TRUE, message=FALSE}
library(plotly)
ggplotly(p)
```


## References 

