---
title: Functional Enrichment Analysis 
author: "First/last name (first.last@ucr.edu)"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`" 
output:
  html_document:
    toc: true
    toc_float:
        collapsed: true
        smooth_scroll: true
    toc_depth: 3
    fig_caption: yes
    code_folding: show
    number_sections: true

fontsize: 14pt
bibliography: bibtex.bib
weight: 11
type: docs
---

<!--
- Compile from command-line
Rscript -e "rmarkdown::render('rfea.Rmd', c('html_document'), clean=F); knitr::knit('rfea.Rmd', tangle=TRUE)"
-->

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
options(width=100, max.print=1000)
knitr::opts_chunk$set(
    eval=as.logical(Sys.getenv("KNITR_EVAL", "TRUE")),
    cache=as.logical(Sys.getenv("KNITR_CACHE", "TRUE")))
```

```{r setup, echo=FALSE, messages=FALSE, warnings=FALSE}
suppressPackageStartupMessages({
    library(ggplot2)
    library(fgsea)
})
```

<div style="text-align: right"> 
Source code downloads: &nbsp; &nbsp;
[ [.Rmd](https://raw.githubusercontent.com/tgirke/GEN242//main/content/en/tutorials/rfea/rfea.Rmd) ] &nbsp; &nbsp; 
[ [.R](https://raw.githubusercontent.com/tgirke/GEN242//main/content/en/tutorials/rfea/rfea.R) ]
</div>


## Introduction

The following introduces several widely used gene and protein annotation
systems that are commonly used for functional enrichment analysis (FEA). These
include among many other annotation systems: Gene Ontology (GO), Disease
Ontology (DO) and pathway annotations, such as KEGG and Reactome. Examples of
widely used statistical enrichment methods are introduced as well. These
statistical FEA methods assess whether functional annotation terms are
over-represented in a query gene set. In case of so called over-represention
analysis (ORA) methods the query is a list of unranked gene identifiers
[@Falcon2007-eb]. In contrast to this, Gene Set Enrichment Analysis (GSEA)
algorithms use as query score ranked lists, _e.g._ all genes profiled by an assay
[@Subramanian2005-kx; @Sergushichev2016-ms; @Duan2020-wz].  The actual sets can
be composed of genes, proteins and even compounds. For simplicity. the term
gene sets is used throughtout this text.

## Functional Annotations Systems
This section introduces a small selection of functional annotation systems, largely
 provided by Bioconductor packages. This includes code to inspect how the annotations 
are organized and how to access them.

## Gene Ontology DB 

`GO.db` is a data package that stores the GO term information in an SQLite database.
Several accessor functions are provide to query the database. Organism specific
gene to GO annotations are provied by organism data packages and/or Bioconductor's 
[`AnntationHub`](https://bioconductor.org/packages/release/bioc/html/AnnotationHub.html).
The following provide sample code for using `GO.db` as well as a organism database
example.

```{r godb, eval=FALSE, warning=FALSE, message=FALSE}
## Load GOstats library
library(GOstats); library(GO.db)
## Print complete GO term information for "GO:0003700"
GOTERM$"GO:0003700"
## Print parent and children terms for a GO ID
GOMFPARENTS$"GO:0003700"; GOMFCHILDREN$"GO:0003700"
## Print complete lineages of parents and children for a GO ID
GOMFANCESTOR$"GO:0003700"; GOMFOFFSPRING$"GO:0003700"
## Print number of GO terms in each of the 3 ontologies
zz <- eapply(GOTERM, function(x) x@Ontology); table(unlist(zz))
## Gene to GO mappings for an organism (here Arabidopsis)
library(org.At.tair.db) # For human use org.Hs.eg.db
xx <- as.list(org.At.tairGO2ALLTAIRS)
```

## Pathways 

### KEGG

#### `KEGG.db`

### Reactome

#### `reactome.db`

#### `ReactomeContentService4R`

## Functional Enrichment Analysis Methods

### Over-representation analysis (ORA) 

#### `GOstats` Package

```{r gostats, eval=FALSE, warning=FALSE, message=FALSE}
## Load required packages
library(GOstats); library(GO.db); library(org.At.tair.db)
## Define universe and test sample set
geneUniverse <- keys(org.At.tairGENENAME)
geneSample <- c("AT2G46210", "AT2G19880", "AT2G38910", "AT5G25140", "AT2G44525")
## Generate params object
params <- new("GOHyperGParams", geneIds = geneSample,
                universeGeneIds = geneUniverse,
                annotation="org.At.tair", ontology = "MF", pvalueCutoff = 0.5,
                conditional = FALSE, testDirection = "over")
## Run enrichment test
hgOver <- hyperGTest(params)
## Viewing of results
summary(hgOver)[1:4,]
htmlReport(hgOver, file = "MyhyperGresult.html")
```

#### `GOHyperGAll` and `GOCluster_Report`
```{r gohypergall, eval=FALSE, warning=FALSE, message=FALSE}
library(systemPipeR)
## Obtain annotations from BioMart
listMarts() # To choose BioMart database
m <- useMart("ENSEMBL_MART_PLANT"); listDatasets(m) 
m <- useMart("ENSEMBL_MART_PLANT", dataset="athaliana_eg_gene")
listAttributes(m) # Choose data types you want to download
go <- getBM(attributes=c("go_accession", "tair_locus", "go_namespace_1003"), mart=m)
go <- go[go[,3]!="",]; go[,3] <- as.character(go[,3])
write.table(go, "GOannotationsBiomart_mod.txt", quote=FALSE, row.names=FALSE, col.names=FALSE, sep="\t")

## Create catDB instance (takes a while but needs to be done only once)
catdb <- makeCATdb(myfile="GOannotationsBiomart_mod.txt", lib=NULL, org="", colno=c(1,2,3), idconv=NULL)
catdb

## Create catDB from Bioconductor annotation package
# catdb <- makeCATdb(myfile=NULL, lib="ath1121501.db", org="", colno=c(1,2,3), idconv=NULL)

## AffyID-to-GeneID mappings when working with AffyIDs 
# affy2locusDF <- systemPipeR:::.AffyID2GeneID(map = "ftp://ftp.arabidopsis.org/home/tair/Microarrays/Affymetrix/affy_ATH1_array_elements-2010-12-20.txt", download=TRUE)
# catdb_conv <- makeCATdb(myfile="GOannotationsBiomart_mod.txt", lib=NULL, org="", colno=c(1,2,3), idconv=list(affy=affy2locusDF))
# systemPipeR:::.AffyID2GeneID(catdb=catdb_conv, affyIDs=c("244901_at", "244902_at"))

## Next time catDB can be loaded from file
save(catdb, file="catdb.RData") 
load("catdb.RData")

## Perform enrichment test on single gene set
test_sample <- unique(as.character(catmap(catdb)$D_MF[1:100,"GeneID"]))
GOHyperGAll(catdb=catdb, gocat="MF", sample=test_sample, Nannot=2)[1:20,]

## GO Slim analysis by subsetting results accordingly
GOHyperGAll_result <- GOHyperGAll(catdb=catdb, gocat="MF", sample=test_sample, Nannot=2)
GOHyperGAll_Subset(catdb, GOHyperGAll_result, sample=test_sample, type="goSlim") 

## Reduce GO term redundancy in 'GOHyperGAll_results'
simplifyDF <- GOHyperGAll_Simplify(GOHyperGAll_result, gocat="MF", cutoff=0.001, correct=T)
# Returns the redundancy reduced data set. 
data.frame(GOHyperGAll_result[GOHyperGAll_result[,1] 

## Batch Analysis of Gene Clusters
testlist <- list(Set1=test_sample)
GOBatchResult <- GOCluster_Report(catdb=catdb, setlist=testlist, method="all", id_type="gene", CLSZ=10, cutoff=0.001, gocats=c("MF", "BP", "CC"), recordSpecGO=c("GO:0003674", "GO:0008150", "GO:0005575"))

## Plot 'GOBatchResult' as bar plot
goBarplot(GOBatchResult, gocat="MF")
```

### Set enrichment analysis (SEA) 

#### `fgsea` Package


## Version Information

```{r sessionInfo}
sessionInfo()
```

## References
